Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-05 cordillera-cycle.tex
Update author list again.
Completely rework the intro.

% Authors
\Author[1,2]{J.}{Seguinot}
\Author[3]{M.}{Margold}
\Author[2]{I.}{Rogozhina}
\Author[1]{A.~P.}{Stroeven}
\Author[1]{J.}{Kleman}
\runningauthor{J.~Seguinot et~al.}
\correspondence{J.~Seguinot (julien.seguinot@natgeo.su.se)}

@@

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera were more extensive than today. At the last glacial maximum, a
contiguous blanket of ice extended from the Alaska Range in the north to the
North Cascades in the south (Fig.~\ref{fig:locmap}). This ice mass is known as
the former Cordilleran ice sheet \citep{Dawson.1888}.

For more than a century, exploration and geological investigation of the
Cordillera have led to a large collection of evidence of the former ice cover
\citep{Jackson.Clague.1991}. This evidence consists of mapped boundaries of the
former ice extent (\citealp[Fig.~1.12]{Clague.1989}; \citealp{Duk-Rodkin.1999};
\citealp{Dyke.2004}), directions of past sliding velocities
(\citealp{Prest.etal.1968}; \citealp[Fig.~2]{Kleman.etal.2010}), and locations
of former
melt-water streams \citep{Margold.etal.2011, Margold.etal.2013a}, as well as the
timing of glaciation identified from radiocarbon dating \citep[e.g.,][]
{Clague.1981, Porter.Swanson.1998}, cosmogenic dating \citep[e.g.,][]
{Ward.etal.2007, Menounos.etal.2008, Stroeven.etal.2010, Stroeven.etal.2014},
and the offshore sedimentary record \citep{Cosma.etal.2008, Davies.etal.2011}%
\footnote{\textbf{Arjen, Martin,}, please comment on my choice of references
    in this paragraph. Is it a fair sample of the current state of the field?
    Am I missing some other important works?}.

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last
glacial cycle (\citealp[Fig. 1.12]{Clague.1989}; \citealp{Dyke.2004}). However,
former ice thickness
and the ice sheet's contribution to the last glacial maximum sea-level
low-stand remain poorly constrained. Moreover, our understanding of the
glaciation history is mainly restricted to the deglaciation phase and a
regional scale, while little is known about the ice sheet evolution prior to
the last glacial maximum extent \citep[Fig.~6]{Kleman.etal.2010}. Although
time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
patterns are available for the Eurasian and Laurentide and ice sheets
\citep{Kleman.etal.1997, Kleman.etal.2010}, such is not the case for the
Cordilleran ice sheet, where complex arrangements of ice flow directions
emanating from multiple glaciation centres have been identified
\citep[Fig.~1.12]{Prest.etal.1968, Clague.1989}, but remain poorly understood
\citep[p.~2049]{Kleman.etal.2010}.

The present study uses a numerical ice sheet model \citep{PISM-authors.2014},
calibrated against field-based evidence to perform a quantitative
reconstruction of the Cordilleran ice sheet history through the last glacial
cycle. Although numerical modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet more than twenty years
ago (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991}), the
ubiquitously mountainous
topography of the region has presented a major challenge to its application.
In fact, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} allows
high-resolution modelling of palaeo-glaciers on mountainous terrain
over millenial time scales \citep{Golledge.etal.2012}. In addition, complex
topography of the North American Cordillera induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as input to the ice sheet model
\citep{Seguinot.etal.2013}.

Because past climate conditions are subject to considerable uncertainty, our
palaeoclimate forcing is a simplistic approximation of time-evolved temperature
and precipitation fields derived from a combination of a present-day
atmospheric reanalysis \citep{Mesinger.etal.2006}, lapse-rate corrections,
and temperature offset time series. The latter are obtained by scaling six
different palaeo-temperature reconstructions from proxy records around the
globe, including two \chem{\delta^{18}O} records from Greenland ice cores
\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two \chem{\delta^{18}O}
records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores \citep{Herbert.etal.2001}. Model output is compared to
geomorphological evidence in terms of timing and extent of glaciation and
patterns of deglaciation.

@@

The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
near-surface air temperature and precipitation rate fields of the NARR
\citep{Mesinger.etal.2006} over the period 1979--2000. Modern climate of the
North American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability
(Fig.~\ref{fig:atm}).
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as identified in a previous sensitivity
study \citep{Seguinot.etal.2013}.

@@

On the one hand, this could indicate that the model performs poorly in this
region, perhaps because the modelled last glacial maximum ice sheet is too
thick, or too extensive to the west. Because feedback mechanisms between ice
sheet topography and regional climate such as wind redistribution, orographic
precipitation and latent warming of moisture-depleted air are absent in our
model, we regard this possibility as fully probable \citep{Seguinot.etal.2013}.
On the other hand, in both simulations driven by the GRIP and EPICA records,
the deglaciation of central British Columbia consists of a rapid northwards
retreat associated with southwards-flowing, non-sliding ice lobes on the
Interior Plateau, while the surrounding mountain ranges are already ice-free
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
appears compatible with the prevailing conceptual model of deglaciation of
central British Columbia, in which mountain ranges emerge from the ice before
the plateau \citep[Fig.~7]{Fulton.1991}. If it is valid, it may suggest that
glacial lineations on the Interior Plateau of British Columbia may be of older
age than the last glacial maximum, and may have remained intact throughout the
deglaciation.

>>>>

% Authors
\Author[1,2]{J.}{Seguinot}
\Author[2]{I.}{Rogozhina}
\Author[3]{M.}{Margold}
\Author[1]{J.}{Kleman}
\Author[1]{A.~P.}{Stroeven}
\runningauthor{J.~Seguinot et~al.}
\correspondence{J.~Seguinot (julien.seguinot@natgeo.su.se)}

@@

During most of the the last glacial cycle, glaciers and ice caps of the North
American Cordillera were more extensive than today. At the Last Glacial
Maximum (LGM), a continuous blanket of ice, the Cordilleran ice sheet
\citep{Dawson.1888}, extended from the Alaska Range in the north to the
North Cascades in the south. It calved into the Pacific Ocean in the west, and
merged with the Laurentide ice sheet in the east (Fig.~\ref{fig:locmap}).

For more than a century, exploration and geological investigations of the
Cordilleran mountains have led to many observations in support of the former
ice sheet
    \citep{Jackson.Clague.1991}.
Despite the lack of documented end moraines offshore, in the zone of confluence
with the Laurentide ice sheet, and in areas swept by the Missoula floods
    \citep{Carrara.etal.1996},
the south-western and north-eastern margins provide key constraints that
allow reasonable reconstructions of the maximum ice extent
    (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
     \citealp{Duk-Rodkin.1999};
     \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
The timing of the last glaciation, reconstructed from radiocarbon dating
    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
           Menounos.etal.2008},
cosmogenic exposure dating
    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.inpress},
bedrock deformation in response to the ice load
    \citep{Clague.James.2002, Clague.etal.2005},
and the offshore sedimentary record
    \citep{Cosma.etal.2008, Davies.etal.2011}
indicate that the LGM ice-sheet was short-lived. However, former ice thickness
and, therefore, the ice sheet's contribution to the last glacial maximum sea
level low stand
    \citep{Carlson.Clark.2012, Clark.Mix.2002}
remain uncertain.

Our understanding of the Cordilleran glaciation history prior to the LGM is
even more fragmentary.
In parts of the Yukon Territory and Alaska, dated glacial erratics indicate
an extensive Marine Isotope Stage (MIS)~4 glaciation
    \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
           Stroeven.etal.2010, Stroeven.etal.2014},
yet it is not known whether other regions were affected. Lineation patterns in
the ice sheet interior include flow directions likely older than the LGM
    \citep[Fig.~2]{Kleman.etal.2010},
but their absolute age is uncertain.

In contrast, our understanding of the deglaciation history of the Cordilleran
ice sheet since the LGM is considerable, albeit mostly at a regional scale.
Geomorphological evidence from south-central British Columbia suggests a rapid
deglaciation, including early emergence of the high-elevated areas and
stagnation of thin ice cover in the lowlands
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a},
yet this model may not apply in all areas of the Cordilleran ice sheet
    \citep{Margold.etal.2013}.
Finally, disparate clues for a late-glacial re-advance have been found in the
Coast, Columbia and Rocky Mountains
    \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
           Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
           Menounos.etal.2008},
but the evidence is much more sparse than that of regions surrounding the North
Atlantic \aref, thus leaving space for debate around the possibility of a
regional late-glacial cold reversal.

In general, the topographic complexity of the Cordillera has not allowed for
ice-sheet wide reconstructions of glacial advance and retreat patterns such as
those available for the Fennoscandian and Laurentide and ice sheets
     \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
            Kleman.etal.1997, Kleman.etal.2010}.
Here, we use a numerical ice sheet model \citep{PISM-authors.2014},
calibrated against field-based evidence, to perform a quantitative
reconstruction of the ice sheet history through the last glacial cycle, and
analyse some of the long-standing questions related to its evolution:

\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the
    last glacial maximum?
  \item Which were the primary dispersal centres? Do they reflect a stable or
    ephemeral configuration?
  \item What was the scale of glaciation prior to the last glacial maximum?
  \item How rapid was the deglaciation? Did it include stand-stills or
    re-advances?
\end{itemize}

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet more than twenty years
ago (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000})
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} allows
high-resolution numerical modelling of palaeo-glaciers on mountainous terrain
over millenial time scales \citep{Golledge.etal.2012}. Second, the complex
topography of the North American Cordillera also induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as input to the ice sheet model
\citep{Seguinot.etal.2014}.

Because climate conditions over the last glacial cycle are subject to
considerable uncertainty, our palaeoclimate forcing is a simplistic approach
including time-evolved temperature and precipitation fields derived from a
combination of a present-day atmospheric reanalysis \citep{Mesinger.etal.2006},
lapse-rate corrections,
and temperature offset time series. The latter are obtained by scaling six
different palaeo-temperature reconstructions from proxy records around the
globe, including two oxygen isotopes records from Greenland ice cores
\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two oxygen isotopes
records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores \citep{Herbert.etal.2001}. Model output is compared to
geomorphological evidence in terms of timing and extent of glaciation and
patterns of deglaciation.

@@

The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
near-surface air temperature and precipitation rate fields of the NARR
\citep{Mesinger.etal.2006} over the period 1979--2000. Modern climate of the
North American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability
(Fig.~\ref{fig:atm}).
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as identified in a previous sensitivity
study \citep{Seguinot.etal.2014}.

@@

On the one hand, this could indicate that the model performs poorly in this
region, perhaps because the modelled last glacial maximum ice sheet is too
thick, or too extensive to the west. Because feedback mechanisms between ice
sheet topography and regional climate such as wind redistribution, orographic
precipitation and latent warming of moisture-depleted air are absent in our
model, we regard this possibility as fully probable \citep{Seguinot.etal.2014}.
On the other hand, in both simulations driven by the GRIP and EPICA records,
the deglaciation of central British Columbia consists of a rapid northwards
retreat associated with southwards-flowing, non-sliding ice lobes on the
Interior Plateau, while the surrounding mountain ranges are already ice-free
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
appears compatible with the prevailing conceptual model of deglaciation of
central British Columbia, in which mountain ranges emerge from the ice before
the plateau \citep[Fig.~7]{Fulton.1991}. If it is valid, it may suggest that
glacial lineations on the Interior Plateau of British Columbia may be of older
age than the last glacial maximum, and may have remained intact throughout the
deglaciation.
